How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-15 talk/dyla2007/dyla.tex
More clarifications based on reviewers' feedback.

\usepackage{makeidx}
\usepackage{graphicx}

@@

Dynamic languages are traditionally implemented by writing a virtual
machine for them in a low-level language like C or in a language that
can relatively easily be turned into C.  The machine implements an
object model supporting the high level dynamic language's objects.  It
typically provides features like automatic garbage collection.  Recent
languages like Python, Ruby, Perl and JavaScript have complicated
semantics which are most easily mapped to a simple interpreter operating
on syntax trees or bytecode; simpler languages like Lisp and Self
typically have more efficient implementations based on code
generation.

@@

PyPy contains a Python interpreter implemented in Python, from which
Python VMs can be generated.  The reader is referred to
\cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
again that the argument we make in the present paper is \emph{not} that
VMs for dynamic languages should be written in their own host language
(as many projects like Squeak \cite{Squeak} and others have done) but
instead that VMs should not be \emph{written} in the first place -- they
should be generated from simple interpreters written in any suitable
high-level language.

@@

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a
higher memory overhead to start with.  For example, an instance of Sun's
Java VM which just loaded Jython consumes XXX MB of non-shared memory
(XXX for the JVM and an additional XXX after loading Jython), while a
CPython process fits in XXX MB.

@@

Another important point that makes implementation of languages on top of OO VMs
harder is that typically OO VMs don't support meta-programming very well, or do so only
at the bytecode level.
\end{itemize}

@@

The deeper problem with the otherwise highly-tuned JIT compilers of the
OO VMs is that they are not a very good match for running dynamic
languages.  It might be possible to tune a general-purpose JIT compiler
enough and write the dynamic language implementation accordingly so
that most of the bookkeeping work involved in running the dynamic
language can be removed -- dispatching, boxing, unboxing...  However
this has not been demonstrated yet.

By far the fastest Python implementation, Psyco \cite{psyco-software}, contains a
hand-written language-specific dynamic compiler.  PyPy's translation
tool-chain is able to extend the generated VMs with an automatically
generated dynamic compiler that uses techniques similar to those of Psyco
\cite{Psyco-paper}, derived from the
interpreter.  This is achieved by a pragmatic application of partial
evaluation techniques guided by a few hints added to the source of the
interpreter.  In other words, it is possible to produce a reasonably
good language-specific JIT compiler and insert it into a VM, alongside
with the necessary support code and the rest of the regular interpreter.

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{Do not write dynamic language implementations ``by hand''.}
Writing them more abstractly, at a higher level, has primarily only
advantages, among them the avoidance of a proliferation of diverging
implementations.  Writing interpreters both flexibly and efficiently
is difficult and meta-programming is a good way to achieve it.
Moreover, this is not incompatible with targeting and benefiting from
existing high-quality object-oriented virtual machines like those of the Java and .NET.

\item \emph{Do not write VMs ``by hand''.}
Writing language-specific virtual machines is a time-consuming task for
medium to large languages.  Unless large amounts of resources can be
invested, the resulting VMs are bound to have limitations which lead to
the emergence of many implementations, a fact that is taxing precisely
for a community with limited resources.  (This is of course even more
true for general-purpose VMs.)
\end{itemize}

@@

\begin{itemize}
\item \emph{Let's write more meta-programming translation toolchains.}
Aside from the advantages described in section
\ref{sect:metaprogramming}, a translation toolchain need not be
standardized for inter-operability but can be tailored to the needs of
each project.  Diversity is good; there is no need to attempt to
standardize on a single OO VM.
\end{itemize}

>>>>

\usepackage{makeidx}
\usepackage{graphicx}
\sloppy

@@

Dynamic languages are traditionally implemented by writing a virtual
machine for them in a low-level language like C or in a language that
can relatively easily be turned into C.  The machine implements an
object model supporting the high level dynamic language's objects.  It
typically provides features like automatic garbage collection.  Recent
languages like Python, Ruby, Perl and JavaScript have complicated
semantics which are most easily mapped to a naive interpreter operating
on syntax trees or bytecode; simpler languages\footnote
{
In the sense of the primitive semantics.  ``Simple'' here is
as opposed to ``complicated'', not as opposed to ``complex'': Common
Lisp for example is not a small language, but it can at least in theory
be expressed from a smaller core of primitives.  In Python, all
primitive operations have complicated semantics.  The argument developed
in the present paper is more relevant to ``complicated'' dynamic languages.
}
like Lisp, Smalltalk and Self typically have more
efficient implementations based on code generation.

@@

PyPy contains a Python interpreter implemented in Python, from which
Python VMs can be generated.  The reader is referred to
\cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
again that the argument we make in the present paper is \emph{not} that
VMs for dynamic languages should be written in their own host language
(as many projects like Squeak \cite{Squeak} and others have done) but
instead that VMs should not be \emph{written} in the first place -- they
should be generated from simple interpreters written in any suitable
high-level\footnote{``High-level'' is taken by opposition to languages
like Scheme48's PreScheme \cite{kelsey-prescheme} or Squeak's \cite{Squeak}
SLang which use the syntax and
metaprogramming facilities of a high-level language but encode
low-level details like object layout and memory management.} language.

@@

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a
higher memory overhead to start with.  For example, an instance of Sun's
Java VM which just loaded Jython consumes 34-42 MB of memory, while a
CPython process fits in 3-4 MB.

@@

Another important point that makes implementation of languages on top of OO VMs
harder is that typically general-purpose OO VMs don't support meta-programming
very well, or do so only at the bytecode level.
\end{itemize}

@@

The deeper problem with the otherwise highly-tuned JIT compilers of the
OO VMs is that they are not a very good match for running dynamic
languages.  It might be possible to tune a general-purpose JIT compiler
enough and write the dynamic language implementation accordingly so
that most of the bookkeeping work involved in running the dynamic
language can be removed -- dispatching, boxing, unboxing...  However
this has not been demonstrated yet.\footnote
{Still in the draft stage, a proposed
extension to the Java bytecode \cite{invokedynamic} might help achieve
better integration between the Java JITs and dynamic language
implementations running on top of JVMs.}

By far the fastest Python implementation, Psyco \cite{psyco-software}
contains a hand-written language-specific dynamic compiler.  It works by
specializing (parts of) Python functions by feeding runtime information
back into the compiler (typically, but not exclusively, object types).
The reader is referred to \cite{Psyco-paper} for more details.

PyPy abstracts on this approach: its translation tool-chain is able to
extend the generated VMs with an \emph{automatically generated} dynamic
compiler that uses techniques similar to those of Psyco, derived from
the interpreter.  This is achieved by a pragmatic application of partial
evaluation techniques guided by a few hints added to the source of the
interpreter.  In other words, it is possible to produce a reasonably
good language-specific JIT compiler and insert it into a VM, alongside
with the necessary support code and the rest of the regular interpreter.

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{High-level languages are suitable to implement dynamic languages.}
They allow an interpreter to be written more abstractly, which has many
advantages -- among them the avoidance of a proliferation of diverging
implementations, and better ways to combine flexibility with efficiency.
Moreover, this is not incompatible with targeting and benefiting from
existing high-quality object-oriented virtual machines like those of the
Java and .NET.

\item \emph{Do not write VMs ``by hand''.}
In other words, write an \emph{interpreter} but not a
\emph{virtual machine} for the language.
Writing language-specific virtual machines is a time-consuming task for
medium to large languages.  Unless large amounts of resources can be
invested, the resulting VMs are bound to have limitations which lead to
the emergence of many implementations, a fact that is taxing precisely
for a community with limited resources.  (This is of course even more
true for general-purpose VMs.)
\end{itemize}

@@

\begin{itemize}
\item \emph{Let's write more meta-programming translation toolchains.}
Aside from the advantages described in section
\ref{sect:metaprogramming}, a translation toolchain need not be
standardized for inter-operability but can be tailored to the needs of
each project.  Diversity is good; translation toolchains offset the need
to attempt to standardize on a single OO VM.
\end{itemize}
